# Figure Construction and Composition

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot)
```

## still need to change column names in res_lassoCV_all_models.csv to lowercase_underscore format for Fold, Dataset, Model, Source, Split

```{r}
# lassoCV full results
full_res <- read_csv("../experiments/compiled_results/res_lassoCV_all_models.csv")
full_res
```

```{r}
# Group and average
cols <- c("rho_score_test", "R2_score_test", "rho_score_train", "R2_score_train", "prot_length")

full_res_avg <- full_res %>%
  group_by(Model, Source, Dataset, Split) %>%
  summarise(across(all_of(cols), mean)) %>%
  mutate(Source = factor(Source, levels = c("viral", "cellular")))

full_res_avg
```

# Figure 1
```{r}
esm2_pooled_avg <- full_res_avg[full_res_avg$Model == "ESM-2 650M" & full_res_avg$Split == "Pooled Split",]
esm2_pooled_avg
```


```{r}
# Figure 1 A
p1a <- ggplot(esm2_pooled_avg, aes(x = prot_length, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  
  scale_x_log10(
    name=expression(italic(R)^2)
  ) +
  
  scale_y_continuous(
    name="Protein Length",
    limits = c(-0.3, 1),
    breaks = c(0, 0.3, 0.6, 0.9)
  ) +

  theme_bw() +
  theme(legend.position = 'none', legend.title = element_blank(), panel.grid.minor = element_blank())

print(p1a)
```




```{r}
# Figure 1 B
p1b <- ggplot(esm2_pooled_avg, aes(x = R2_score_test)) +
  geom_density(aes(fill = Source),  alpha = 0.55, linewidth = 0.25) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  
  scale_y_continuous(name="Density",
                     expand = c(0, 0, 0.05, 0)) +
  
  scale_x_continuous(name=expression(italic(R)^2),
                     expand = c(0, 0, 0.0, 0.0)) +
  
  theme_bw() +
  theme(legend.position = 'none',
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(0.068, 0.068, 0.038, 0.068, "in"))

print(p1b)
```




```{r}
# Figure 1 C
full_pooled_avg <- full_res_avg[full_res_avg$Split == "Pooled Split",] %>% 
 mutate(Model = factor(Model, levels = c('ESM-2 650M', 'ESM C 600M', 'ESM-2-CRVDB', 'ESM-2-URVDB', 'ESM-2 3B Sawhney')))

p1c <- ggplot(full_pooled_avg, aes(x = Model, y = R2_score_test, fill = Source)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.35, alpha = 0.55, linewidth = 0.25) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  
  scale_x_discrete(name='Model') +
  scale_y_continuous(name=expression(italic(R)^2)) +
  
  theme_bw() +

   theme(
    legend.position = 'none',      
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
) 

p1c 



# Figure 1 C - version 2
# full_pooled_avg <- full_res_avg[full_res_avg$Split == "Pooled Split",] %>% 
#  mutate(Model = factor(Model, levels = c('ESM-2 650M', 'ESM C 600M', 'ESM-2-CRVDB', 'ESM-2-URVDB', 'ESM-2 3B Sawhney')))
# 
# p1c <- ggplot(full_pooled_avg, aes(x = Model, y = R2_score_test, fill = Source)) +
#   geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.35, alpha = 0.55, linewidth = 0.25, key_glyph='rect') +
#   scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
#   geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
#   
#   scale_x_discrete(name='Model') +
#   scale_y_continuous(name=expression(italic(R)^2)) +
#   
#   theme_bw() +
# 
#    theme(
#     legend.position = c(1, 1.2),      
#     legend.justification = c("right", "top"),
#     legend.direction = "horizontal",
#     legend.title = element_blank(),
#     panel.grid.major.x = element_blank(),
#     panel.grid.minor.y = element_blank(),
#     plot.margin = margin(t = 50, r = 9, b = 9, l = 9),
# ) 
# 
#   
# p1c 
```





## Composition
```{r,fig.width = 8, fig.height = 7}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = c("right", "top"),
    legend.direction = "horizontal"
  )
)


# Specify positioning for figure composition
fig1_top <- plot_grid(p1a, p1b, ncol = 2, labels = c('A', 'B'), label_fontface = "plain")
fig1_bottom <- plot_grid(p1c, labels = c('C'), label_fontface = "plain")

# Compose figure 1
fig1 <- plot_grid(
  fig1_top,
  shared_legend,
  fig1_bottom,
  ncol = 1,
  rel_heights = c(1, 0.15, 1)
)

print(fig1)

#ggsave(path = '../experiments/figures', 'comp1_pLMs_res_v03.png', width = 8, height = 7, units = 'in', dpi = 600, bg='white')
```







# Figure 2

```{r}
# Figure 2 A
lassoCV_AVG <- read_csv("../experiments/compiled_results/res_lassoCV_x_AVG.csv", show_col_types = FALSE)

lassoCV_AVG
```


```{r}
p2a <- ggplot(lassoCV_AVG, aes(x = R2_AVG, y = R2_ESM2_650M, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  geom_abline(intercept = 0, alpha = 0.5, linetype = 'dashed') +
  
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  
  scale_x_continuous(
    name = expression(paste(italic(R)^{2}, ' site-means')),
    limits = c(-0.4, 0.85)
      ) +
  scale_y_continuous(
    name = expression(paste(italic(R)^{2}, ' transfer learning')),
    limits = c(-0.4, 0.85)
      ) 
  


print(p2a)
```


```{r}
# Figure 2 B
res_OLS <- read_csv("../experiments/compiled_results/res_OLS.csv", show_col_types = FALSE)
res_OLS
```

```{r}
p2b <- ggplot(res_OLS, aes(x = Model, y = R2, fill = Source)) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, alpha = 0.55, linewidth = 0.25) +
  theme_bw() +
  labs(
    x = 'Independent variable',
    y = expression(paste(italic(R)^2, ' ordinary least squares (OLS)'))
    ) +
  
  theme(
    legend.position = 'none',
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
) 
  
print(p2b)
```

## Composition
```{r, fig.width=8, fig.height=4}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = c("right", "top"),
    legend.direction = "horizontal"
  )
)

# Compose figure 2
fig2 <- plot_grid(
  shared_legend,
  plot_grid(p2a, p2b, ncol = 2, labels = c('A', 'B'), 
            label_fontface = "plain", rel_widths = c(1, 1)),
  ncol = 1,
  rel_heights = c(0.1, 1)
)


print(fig2)
#ggsave(path='../experiments/figures/', filename='comp2_lassoCV_AVG_OLS_v03.png', width = 8, height = 4, units = 'in', dpi = 600, bg='white')
```


# Figure 3
```{r}
# Load variability metrics results
res_var <- read_csv("../experiments/compiled_results/res_lassoCV_esm2_650m_all_metrics.csv", show_col_types = FALSE)
res_var <- res_var[res_var$Split == "Pooled Split",]
res_var
```


```{r}
# Figure 3 A
corr_viral <- cor.test(res_var[res_var$Source == "viral",]$RVSM, res_var[res_var$Source == "viral",]$R2_score_test)
corr_cellular <- cor.test(res_var[res_var$Source == "cellular",]$RVSM, res_var[res_var$Source == "cellular",]$R2_score_test)

p3a <- ggplot(res_var, aes(x = RVSM, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  annotate(
    "text",
    x = 0.401, y = 0.82,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_viral$estimate),
    color = '#D55E00B3'
  ) +
  annotate(
    "text",
    x = 0.401, y = 0.76,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_cellular$estimate),
    color = '#0072B2B3'
  ) +
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  
  labs(
    y = expression(paste(italic(R)^2, ' transfer learning')),
    x = "Relative variability of site means"
  )

print(p3a)
```



```{r}
# Figure 3 B
corr_RF_viral <- cor.test(res_var[res_var$Source == "viral",]$RVSM, res_var[res_var$Source == "viral",]$FHVS)
corr_RF_cellular <- cor.test(res_var[res_var$Source == "cellular",]$RVSM, res_var[res_var$Source == "cellular",]$FHVS)

p3b <- ggplot(res_var, aes(x = RVSM, y = FHVS, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  #geom_smooth(method = 'lm', se = FALSE) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  annotate(
    "text",
    x = 0.7, y = 0.90,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_RF_viral$estimate),
    color = '#D55E00B3'
  ) +
  annotate(
    "text",
    x = 0.7, y = 0.85,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_RF_cellular$estimate),
    color = '#0072B2B3'
  ) +
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  labs(y = 'Fraction of highly variable sites', x = 'Relative variability of site means')

print(p3b)
```

```{r}
# Figure 3 C -- unfinished
p3c <- ggplot(res_var, aes(x = FHVS, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_smooth(method = 'lm', formula = 'y ~ I(x*x) + x', se=FALSE, size=0.7, color = '#63636380') +

  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  labs(y = expression(paste(italic(R)^2, " transfer learning")),
        x = "Fraction of highly variable sites"
  )
  
print(p3c)
```

```{r}
# Figure 3 D
lassoCV_AVG <- read_csv("../experiments/compiled_results/lassoCV_x_AVG_all.csv", show_col_types = FALSE)
lassoCV_AVG

p3d <- ggplot(lassoCV_AVG, aes(x = R2_AVG_HVS, y = R2_ESM2_650M_HVS, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke = 0) +
  geom_smooth(method = 'lm', se = FALSE) +
  geom_abline(intercept = 0, alpha = 0.5, linetype = 'dashed') +
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  labs(
    y = expression(paste(R^{2}, ' transfer learning')),
    x = expression(paste(R^{2}, ' site-means'))) +
  xlim(-1.00, 0.82) +
  ylim(-1.00, 0.82)

print(p3d)
```

## Composition

```{r}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = "center",
    legend.direction = "horizontal"
  )
)

# Compose figure 3
fig3 <- plot_grid(
  shared_legend,
  plot_grid(p3a, p3b, p3c, p3d, ncol = 2, nrow = 2, labels = c('A', 'B', 'C', 'D'), rel_widths = c(1, 1), rel_heights = c(1, 1)),
  ncol = 1,
  rel_heights = c(0.07, 1)
)

print(fig3)

ggsave('comp3_site_variation4_v03.png', width = 8, height = 7, units = 'in', dpi = 300, path = '../experiments')
```


# Figure 4
```{r}
# Figure 4
p4 <- ggplot(full_res_avg, aes(x = Model, y = R2_score_test, fill = Split)) +
  scale_fill_manual(values = c("Pooled Split" = "#20B15C", "Site Split" = "#E2B40E"),
                    labels = c("Pooled", "Stratified by Site")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.5, alpha = 0.5) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  scale_x_discrete(labels = c("ESM-2 650M", "ESM C 600M", "ESM-2 650M CRVDB", "ESM-2 650M URVDB", "ESM-2 3B Sawhney")) +
  labs(x = 'Model', y = expression(R^{2})) +
  theme_bw() +
  facet_wrap(~Source, nrow = 2) +
  theme(panel.grid.major.x = element_blank(), legend.position = 'top', legend.title = element_blank())

print(p4)

ggsave('fig4_site_x_pooled_all_models_v03.png', width = 8, height = 5, units = 'in', dpi = 300, path = '../experiments')
```

# Figure 5

```{r}
# ProteinGym metrics
prot_gym <- read_csv("../experiments/compiled_results/protein_gym_plus_var_res.csv", show_col_types = FALSE)
kermut <- prot_gym[prot_gym$model_name == "Kermut" & prot_gym$fold_variable_name == "fold_random_5",]
kermut

```

```{r}
model <- lm(Spearman ~ RVSM + FHVS, data = kermut)
summary(model)

# R^2
r2 <- summary(model)$r.squared
print(r2)
```

```{r}
# Figure 5
kermut$y_pred <- predict(model, newdata = kermut)
kermut$y_true <- kermut$Spearman

p5 <- ggplot(kermut, aes(x = y_pred, y = y_true)) +
  geom_point(size = 2.5, alpha = 0.5) +
  geom_smooth(method='lm', se=FALSE, color = 'black') +
  theme_bw() +
  annotate(
    "text",
    x = 0.26, y = 0.91,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("R^2 == %.3f", r2),
    parse = TRUE
  ) +
  labs(x = "Spearman (Predicted)", y = "Spearman (Actual)") +
  theme(legend.position = "top", panel.grid.minor = element_blank())

print(p5)

ggsave('fig5_actual_x_predicted_protGym_v03.png', width = 4, height = 3.5, units = 'in', dpi = 300, path = '../experiments')
```



# Supplementary Figures

## Supplementary 1

## Supplementary 2

## Supplementary 3

## Supplementary 4

```{r}
set.seed(55)

# make axes with ellipses
aa <- c("A", "C", "...", "... ", "W", "Y")
pos_labels <- c("1", "2", "3", "...", " ", "  ", "... ", "L-2", "L-1", "L")

n_aa <- length(aa)
L <- length(pos_labels)

# grid
df <- expand.grid(
  pos = factor(pos_labels, levels = pos_labels),
  aa  = factor(aa, levels = aa)
)

# dataframe for pooled
test_frac <- 0.2
df_rand <- df %>%
  mutate(
    split = ifelse(runif(n()) < test_frac, "test data", "train data"),
    scheme = "Pooled"
  )

# dataframe for stratified by site
test_pos <- sample(pos_labels, size = 2)
df_pos <- df %>%
  mutate(
    split = ifelse(pos %in% test_pos, "test data", "train data"),
    scheme = "Stratified by site"
  )

df_all <- bind_rows(df_rand, df_pos)

# plot
s4 <- ggplot(df_all, aes(x = pos, y = aa, fill = split)) +
  geom_tile(color = "black", linewidth = 0.3, linetype = 'dotted') +
  scale_fill_manual(values = c("train data" = "#e1f7d5", "test data" = "#f1cbff")) +
  facet_wrap(~ scheme, nrow = 1) +
  labs(
    x = "Sequence position index",
    y = "Amino acid substitution"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
    legend.title = element_blank()
  )

print(s4)

ggsave('supp_pooled_site_schematic.png', width = 8, height = 2.5, units = 'in', dpi = 300, path = '../experiments')
```

## Supplementary 5

## Supplementary 6

## Supplementary 7

## did we stil want this as a supplementary figure? pooled v stratified by site performance (Spearman)

```{r}
p4supp <- ggplot(full_res_avg, aes(x = Model, y = rho_score_test, fill = Split)) +
  scale_fill_manual(values = c("Pooled Split" = "#20B15C", "Site Split" = "#E2B40E"),
                    labels = c("Pooled", "Stratified by Site")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.5, alpha = 0.5) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  scale_x_discrete(labels = c("ESM-2 650M", "ESM C 600M", "ESM-2 650M CRVDB", "ESM-2 650M URVDB", "ESM-2 3B Sawhney")) +
  labs(x = 'Model', y = expression(rho)) +
  theme_bw() +
  facet_wrap(~Source, nrow = 2) +
  theme(panel.grid.major.x = element_blank(), legend.position = 'top', legend.title = element_blank())

print(p4supp)

#ggsave('supp4_site_x_pooled_spearman.png', width = 8, height = 5, units = 'in', dpi = 300, path = '../experiments')
```



















