# Figure Construction and Composition

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot)
```

## still need to change column names in res_lassoCV_all_models.csv to lowercase_underscore format for Fold, Dataset, Model, Source, Split

```{r}
# lassoCV full results
full_res <- read_csv("../experiments/compiled_results/res_lassoCV_all_models.csv")
full_res
```

```{r}
# Group and average
cols <- c("rho_score_test", "R2_score_test", "rho_score_train", "R2_score_train", "prot_length")

full_res_avg <- full_res %>%
  group_by(Model, Source, Dataset, Split) %>%
  summarise(across(all_of(cols), mean)) %>%
  mutate(Source = factor(Source, levels = c("viral", "cellular")))

full_res_avg
```




# Figure 1
```{r}
esm2_pooled_avg <- full_res_avg[full_res_avg$Model == "ESM-2 650M" & full_res_avg$Split == "Pooled Split",]
esm2_pooled_avg
```



```{r}
# Figure 1 A
p1a <- ggplot(esm2_pooled_avg, aes(x = prot_length, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  
  scale_x_log10(
    name="Protein length",
    
  ) +
  
  scale_y_continuous(
    name=expression(italic(R)^2),
    limits = c(-0.3, 1),
    breaks = c(0, 0.3, 0.6, 0.9)
  ) +

  theme_bw() +
  theme(legend.position = 'none', legend.title = element_blank(), panel.grid.minor = element_blank())

print(p1a)
```




```{r}
# Figure 1 B
p1b <- ggplot(esm2_pooled_avg, aes(x = R2_score_test)) +
  geom_density(aes(fill = Source),  alpha = 0.55, linewidth = 0.25) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  
  scale_y_continuous(name="Density",
                     expand = c(0, 0, 0.05, 0)) +
  
  scale_x_continuous(name=expression(italic(R)^2),
                     expand = c(0, 0, 0.0, 0.0)) +
  
  theme_bw() +
  theme(legend.position = 'none',
        legend.title = element_blank(),
        panel.grid.minor = element_blank(),
        plot.margin = margin(0.068, 0.068, 0.038, 0.068, "in"))

print(p1b)
```

```{r}
# T test
t.test(R2_score_test ~ Source, data = esm2_pooled_avg)
```



```{r}
# Figure 1 C
full_pooled_avg <- full_res_avg[full_res_avg$Split == "Pooled Split",] %>% 
 mutate(Model = factor(Model, levels = c('ESM-2 650M', 'ESM C 600M', 'ESM-2-CRVDB', 'ESM-2-URVDB', 'ESM-2 3B Sawhney')))

p1c <- ggplot(full_pooled_avg, aes(x = Model, y = R2_score_test, fill = Source)) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.35, alpha = 0.55, linewidth = 0.25) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  
  scale_x_discrete(name='Model') +
  scale_y_continuous(name=expression(italic(R)^2)) +
  
  theme_bw() +

   theme(
    legend.position = 'none',      
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
) 

p1c 
```



```{r}
# T-test
full_pooled_avg %>%
  group_by(Model) %>%
  summarise(
    mean_viral = mean(R2_score_test[Source == "viral"], na.rm = TRUE),
    mean_cellular = mean(R2_score_test[Source == "cellular"], na.rm = TRUE),
    p_value = t.test(R2_score_test ~ Source)$p.value
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

```

```{r}
sprintf("%.5f", 6.197804e-03)

sqrt((0.107 / 5) + (0.303/6))
```



# T-tes across models base ESM-2 (cellular) and fine-tuned ESM-2 (viral) models.

```{r}
full_pooled_avg %>%
  mutate(Model_Source = interaction(Model, Source, sep = " ")) %>% 
  filter(Model_Source %in% c("ESM-2 650M cellular", "ESM-2 650M viral")) %>%
  t.test(R2_score_test ~ Model_Source, data = .)
```

```{r}
full_pooled_avg %>%
  mutate(Model_Source = interaction(Model, Source, sep = " ")) %>% 
  filter(Model_Source %in% c("ESM-2 650M cellular", "ESM-2-CRVDB viral")) %>%
  t.test(R2_score_test ~ Model_Source, data = .)

```

```{r}
full_pooled_avg %>%
  mutate(Model_Source = interaction(Model, Source, sep = " ")) %>% 
  filter(Model_Source %in% c("ESM-2 650M cellular", "ESM-2-URVDB viral")) %>%
  t.test(R2_score_test ~ Model_Source, data = .)
```

```{r}
full_pooled_avg %>%
  mutate(Model_Source = interaction(Model, Source, sep = " ")) %>% 
  filter(Model_Source %in% c("ESM-2 650M cellular", "ESM-2 3B Sawhney viral")) %>%
  t.test(R2_score_test ~ Model_Source, data = .)
```



## Composition
```{r,fig.width = 8, fig.height = 7}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = c("right", "top"),
    legend.direction = "horizontal"
  )
)


# Specify positioning for figure composition
fig1_top <- plot_grid(p1a, p1b, ncol = 2, labels = c('A', 'B'), label_fontface = "plain")
fig1_bottom <- plot_grid(p1c, labels = c('C'), label_fontface = "plain")

# Compose figure 1
fig1 <- plot_grid(
  fig1_top,
  shared_legend,
  fig1_bottom,
  ncol = 1,
  rel_heights = c(1, 0.15, 1)
)

print(fig1)
#ggsave(path = '../experiments/figures', 'comp1_pLMs_res_v03.png', width = 8, height = 7, units = 'in', dpi = 600, bg='white')
```







# Figure 2

```{r}
# Figure 2 A
lassoCV_AVG <- read_csv("../experiments/compiled_results/res_lassoCV_x_AVG.csv", show_col_types = FALSE)

lassoCV_AVG
```


```{r}
p2a <- ggplot(lassoCV_AVG, aes(x = R2_AVG, y = R2_ESM2_650M, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  geom_abline(intercept = 0, alpha = 0.5, linetype = 'dashed') +
  
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = 'top', 
    panel.grid.minor = element_blank()) +
  
  scale_x_continuous(
    name = expression(paste(italic(R)^{2}, ' site-means')),
    limits = c(-0.4, 0.85)
      ) +
  scale_y_continuous(
    name = expression(paste(italic(R)^{2}, ' transfer learning')),
    limits = c(-0.4, 0.85)
      ) 
  

ggsave(path='../experiments/figures/', filename='fig2_lassoCV_X_AVG.png', width = 4, height = 3.5, units = 'in', dpi = 600, bg='white')
print(p2a)
```


```{r}
# Figure 2 B
res_OLS <- read_csv("../experiments/compiled_results/res_OLS.csv", show_col_types = FALSE)
res_OLS
```

```{r}
p2b <- ggplot(res_OLS, aes(x = Model, y = R2, fill = Source)) +
  scale_fill_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, alpha = 0.55, linewidth = 0.25) +
  theme_bw() +
  labs(
    x = 'Independent variable',
    y = expression(paste(italic(R)^2, ' ordinary least squares (OLS)'))
    ) +
  
  theme(
    legend.position = 'none',
    legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank(),
) 
  
print(p2b)
```

## Composition
```{r, fig.width=8, fig.height=4}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = c("right", "top"),
    legend.direction = "horizontal"
  )
)

# Compose figure 2
fig2 <- plot_grid(
  shared_legend,
  plot_grid(p2a, p2b, ncol = 2, labels = c('A', 'B'), 
            label_fontface = "plain", rel_widths = c(1, 1)),
  ncol = 1,
  rel_heights = c(0.1, 1)
)


print(fig2)
#ggsave(path='../experiments/figures/', filename='comp2_lassoCV_AVG_OLS_v03.png', width = 8, height = 4, units = 'in', dpi = 600, bg='white')
```










# Figure 3
```{r}
# Load variability metrics results
res_var <- read_csv("../experiments/compiled_results/res_lassoCV_esm2_650m_all_metrics.csv", show_col_types = FALSE)
res_var <- res_var[res_var$Split == "Pooled Split",]
res_var
```


```{r}
# Figure 3 A
corr_viral <- cor.test(res_var[res_var$Source == "viral",]$RVSM, res_var[res_var$Source == "viral",]$R2_score_test)
corr_cellular <- cor.test(res_var[res_var$Source == "cellular",]$RVSM, res_var[res_var$Source == "cellular",]$R2_score_test)

p3a <- ggplot(res_var, aes(x = RVSM, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  annotate(
    "text",
    x = 0.401, y = 0.82,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_viral$estimate),
    color = '#D55E00B3'
  ) +
  annotate(
    "text",
    x = 0.401, y = 0.76,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_cellular$estimate),
    color = '#0072B2B3'
  ) +
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  
  labs(
    y = expression(paste(italic(R)^2, ' transfer learning')),
    x = "Relative variability of site means (RVSM)"
  )

print(p3a)
```


```{r}
# Figure 3 B
p3b <- ggplot(res_var, aes(x = FHVS, y = R2_score_test, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  geom_smooth(method = 'lm', formula = 'y ~ I(x*x) + x', se=FALSE, size=0.7, color = '#63636380') +

  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  labs(y = expression(paste(italic(R)^2, " transfer learning")),
        x = "Fraction of highly variable sites (FHVS)"
  )
  
print(p3b)
```




```{r}
# Figure 3 C
corr_RF_viral <- cor.test(res_var[res_var$Source == "viral",]$RVSM, res_var[res_var$Source == "viral",]$FHVS)
corr_RF_cellular <- cor.test(res_var[res_var$Source == "cellular",]$RVSM, res_var[res_var$Source == "cellular",]$FHVS)

p3c <- ggplot(res_var, aes(x = RVSM, y = FHVS, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke=0) +
  #geom_smooth(method = 'lm', se = FALSE) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  annotate(
    "text",
    x = 0.7, y = 0.90,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_RF_viral$estimate),
    color = '#D55E00B3'
  ) +
  annotate(
    "text",
    x = 0.7, y = 0.85,
    hjust = 0, vjust = 1,
    size = 4,
    label = sprintf("r = %.3f", corr_RF_cellular$estimate),
    color = '#0072B2B3'
  ) +
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  labs(y = 'Fraction of highly variable sites (FHVS)', x = 'Relative variability of site means (RVSM)')

print(p3c)
```



```{r}
# Figure 3 D
lassoCV_AVG <- read_csv("../experiments/compiled_results/res_lassoCV_x_AVG_HVS.csv", show_col_types = FALSE)


p3d <- ggplot(lassoCV_AVG, aes(x = R2_AVG_HVS, y = R2_ESM2_650M_HVS, color = Source)) +
  scale_color_manual(values = c("viral" = "#D55E00", "cellular" = "#0072B2")) +
  geom_point(size = 3, alpha = 0.55, stroke = 0) +
  geom_line(stat="smooth",method = "lm", size = 0.8, alpha = 0.55) +
  geom_abline(intercept = 0, alpha = 0.5, linetype = 'dashed') +
  
  theme_bw() +
  theme(legend.position = 'none', panel.grid.minor = element_blank()) +
  scale_y_continuous(
    name=expression(paste(italic(R)^2, ' transfer learning')),
    limits = c(-1, 0.9),
  ) +
  scale_x_continuous(
    name=expression(paste(italic(R)^2, ' site-means')),
    limits=c(-1, 0.9),
  )


print(p3d)
```

## Composition

```{r, fig.width=8, fig.height=8}
# Extract the shared legend from one of the plots
shared_legend <- get_legend(
  p1b + theme(
    legend.position = "top",
    legend.justification = c("right", "top"),
    legend.direction = "horizontal"
  )
)




fig3 <- plot_grid(
  plot_grid(p3a, p3b, ncol = 2, nrow = 1, labels = c('A', 'B'), rel_widths = c(1, 1), rel_heights = c(1, 1), label_fontface = "plain"),
  plot_grid(shared_legend, ncol = 1, nrow = 1, rel_heights = 1), 
  plot_grid(p3c, p3d, ncol = 2, nrow = 1, labels = c('C', 'D'), rel_widths = c(1, 1), rel_heights = c(1, 1), label_fontface = "plain"),
  ncol = 1,
  rel_heights = c(1, 0.1, 1)
)

print(fig3)
ggsave(path = '../experiments/figures/', 'comp3_site_variation4_v05.png', width = 8, height = 8, units = 'in', dpi = 600, bg='white')
```








# Figure 4

```{r, fig.width=8, fig.height=6}
# Figure 4
full_res_avg_df = full_res_avg %>% 
  mutate(Model = factor(Model, levels = c('ESM-2 650M', 'ESM C 600M', 'ESM-2-CRVDB', 'ESM-2-URVDB', 'ESM-2 3B Sawhney')))


p4 <- ggplot(full_res_avg_df, aes(x = Model, y = R2_score_test, fill = Split)) +
  scale_fill_manual(#values = c("Pooled Split" = "#20B15C", "Site Split" = "#E2B40E"),
                    values = c("Pooled Split" = "#50AE5A", "Site Split" = "#EF9708"),
                    labels = c("Pooled", "Stratified by Site")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.5, alpha = 0.55, linewidth = 0.25) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  
  labs(x = 'Model', y = expression(italic(R)^2)) +
  
  facet_wrap(~Source, nrow = 2) +
  
  theme_bw(14) +
  theme(
   legend.title = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.direction = 'horizontal',
    legend.position = "inside",
    legend.justification.inside = c(0.98, 0.015),
     )

print(p4)
ggsave(path = '../experiments/figures/', 'fig4_site_x_pooled_all_models_v05.png', width = 8, height = 6, units = 'in', dpi = 600)
```


```{r}

# T-test
full_res_avg_df %>%
  filter(Source == "cellular") %>% 
  group_by(Model) %>%
  summarise(
    mean_poooled = mean(R2_score_test[Split == "Pooled Split"], na.rm = TRUE),
    mean_site = mean(R2_score_test[Split == "Site Split"], na.rm = TRUE),
    p_value = t.test(R2_score_test ~ Split)$p.value
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )
```
```{r}

# T-test
full_res_avg_df %>%
  filter(Source == "viral") %>% 
  group_by(Model) %>%
  summarise(
    mean_poooled = mean(R2_score_test[Split == "Pooled Split"], na.rm = TRUE),
    mean_site = mean(R2_score_test[Split == "Site Split"], na.rm = TRUE),
    p_value = t.test(R2_score_test ~ Split)$p.value
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )
```




# Figure 5

```{r}
# ProteinGym metrics
prot_gym <- read_csv("../experiments/compiled_results/protein_gym_plus_var_metrics_pred_success.csv", show_col_types = FALSE)


prot_gym <- prot_gym %>%
  mutate(
    model_name = case_when(
    model_name == 'Emb. Aug. ESM1v' ~ 'ESM1v',
    model_name == 'Emb. Aug. MSA Transformer' ~ 'MSA Transformer',
    model_name == 'Emb. Aug. Tranception' ~ 'Tranception',
    TRUE ~ model_name  # keep everything else as-is
    ),
    
    model_name = factor(model_name, levels=c('Kermut', 'ProteinNPT', 'MSA Transformer', 'Tranception', 'ESM1v')),
    
    fold_variable_name = case_when(
    fold_variable_name == 'fold_contiguous_5' ~ 'contiguous',
    fold_variable_name == 'fold_modulo_5' ~ 'modulo',
    fold_variable_name == 'fold_random_5' ~ 'random',
    TRUE ~ model_name  # keep everything else as-is
    ),
  ) %>%
  filter(model_name %in% c('Kermut', 'ProteinNPT', 'MSA Transformer', 'Tranception', 'ESM1v'))


prot_gym
```


```{r, fig.width=8, fig.height=4}
p5a <- ggplot(prot_gym, aes(x=model_name, y=Spearman, fill=fct_reorder(fold_variable_name, Spearman, .desc =TRUE))) +
  geom_boxplot(alpha=.55) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  scale_fill_manual(values = c("random" = "#50AE5A", "modulo"="#EF9708", "contiguous" = "#E2E2E2"),
                    labels = c("Pooled", "Modulo", "Contiguous")
                    ) +
  #scale_fill_discrete_diverging(palette = "Green-Orange") + 
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = 'top',
  ) +
  scale_y_continuous(name=Spearman~rho) +
  scale_x_discrete(name='')


p5a
```




```{r}

stat_data <- prot_gym
stat_data$fold_variable_name <- factor(
  stat_data$fold_variable_name,
  levels = c("random", "modulo", "contiguous")
)


models <- unique(stat_data$model_name)

results_list <- lapply(models, function(m) {
  df_m <- subset(stat_data, model_name == m)
  
  fit <- aov(Spearman ~ fold_variable_name, data = df_m)
  posthoc <- TukeyHSD(fit, "fold_variable_name")
  
  data.frame(
    model = m,
    comparison = rownames(posthoc$fold_variable_name),
    p_adj = posthoc$fold_variable_name[, "p adj"]
  )
})


final_results <- do.call(rbind, results_list) %>%
  mutate(
    significance = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>% 
  select(-p_adj) %>% 
  pivot_wider(
    names_from = model,
    values_from = significance,
   
  )


final_results
```





```{r}
kermut <- prot_gym[prot_gym$model_name == "Kermut" & prot_gym$fold_variable_name == "random",]
kermut

model <- lm(Spearman ~ RVSM + FHVS, data = kermut)
summary(model)

# R^2
r2 <- summary(model)$r.squared
print(r2)
```

```{r, fig.width=5, fig.height=4}
# Figure 5
kermut$y_pred <- predict(model, newdata = kermut)
kermut$y_true <- kermut$Spearman

p5b <- ggplot(kermut, aes(x = y_pred, y = y_true)) +
  geom_smooth(method='lm', se=FALSE, color = '#00000080', linewidth=0.3) +
  #geom_point(size=3, alpha=0.7, color='#26456EFF', stroke=0) +
  geom_point(size=3, alpha=0.7, color='#306A36FF', stroke=0) +
  theme_bw() +
  annotate(
    "text",
    x = 0.26, y = 0.91,
    hjust = 0, vjust = 1,
    size = 4,
    label = paste(expression(italic(R)^2), sprintf("== %.3f", r2)),
    parse = TRUE
  ) +
  labs(x = bquote(Spearman~rho~"(Predicted)"), y = bquote(Spearman~rho~"(ProteinGym)")) +
  theme(legend.position = "top", panel.grid.minor = element_blank())

print(p5b)

#ggsave(path = '../experiments/figures/', 'fig5_actual_x_predicted_protGym_v03.3.png', width = 4, height = 3.5, units = 'in', dpi = 300, )
```



```{r}
cols <- c("Spearman", "R2")

success_df <- prot_gym %>%
  filter(fold_variable_name == "random") %>%
  group_by(model_name) %>%
  summarise(across(all_of(cols), mean), .groups = "drop") %>%
  arrange(desc(Spearman)) %>%
  mutate(model_name = fct_reorder(model_name, Spearman, .desc = TRUE))

success_df 
```

```{r, fig.width=4, fig.height=4}
p5c <- ggplot(success_df, aes(y=Spearman, x=R2, color=fct_reorder(model_name, Spearman))) +
  geom_point(size=3, alpha=1) +
  scale_color_manual(values = c('Kermut'='#2CA02E', 'ProteinNPT'='#2592DB', 'Tranception'='#CF750D', 'MSA Transformer'='#B669E8', 'ESM1v'='#948890')) +
  geom_smooth(method='lm', se=FALSE, color = '#00000080', linewidth=0.3) +
  
  labs(y = bquote('Mean'~Spearman~rho~"(ProteinGym)"), x = bquote(italic(R)^2~"of linear model predicting"~Spearman~rho)) +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position='top',
    legend.key.size = unit(0.05, "in"),
    
    ) 


p5c <- p5c + guides(color = guide_legend(nrow = 2, byrow = TRUE))
p5c
```



```{r, fig.width=8, fig.height=7}
# Specify positioning for figure composition
fig5_top <- plot_grid(p5a, labels = c('A'), label_fontface = "plain")
fig5_bottom <- plot_grid(p5b, p5c, ncol = 2, labels = c('B', 'C'), label_fontface = "plain", rel_widths = c(1, 1))

# Compose figure 1
fig5 <- plot_grid(
  fig5_top,
  fig5_bottom,
  ncol = 1,
  rel_heights = c(1, 1)
)

print(fig5)

ggsave(path = '../experiments/figures/', 'comp5_protGym_v02.4.png', width = 8, height = 7.5, units = 'in', dpi = 600, bg='white')
```







```{r, fig.width=6, fig.height=4}
# library(MASS)
# 
# dens <- kde2d(kermut$y_pred, kermut$y_true, n = 150)
# ix <- findInterval(kermut$y_pred, dens$x)
# iy <- findInterval(kermut$y_true, dens$y)
# kermut$density <- dens$z[cbind(ix, iy)]
# kermut$density <- kermut$density / max(kermut$density) # normilize
# 
# p5 <- ggplot(kermut, aes(x = y_pred, y = y_true)) +
#   geom_smooth(method='lm', se=FALSE, color = '#00000080', size=0.5) +
#   geom_point(aes(color = density), size=2, alpha=0.8, stroke=0) +
#   scale_color_viridis_c(name = "Point density") +
#   theme_bw() +
#   annotate(
#     "text",
#     x = 0.26, y = 0.91,
#     hjust = 0, vjust = 1,
#     size = 4,
#     label = paste(expression(italic(R)^2), sprintf("== %.3f", r2)),
#     parse = TRUE
#   ) +
#   labs(x = "Spearman (Predicted)", y = "Spearman (Actual)") +
#   theme(
#     legend.position = "right", 
#     #legend.direction = "horizontal",
#     panel.grid.minor = element_blank(),
#     legend.title = element_text(size = 10, margin = margin(b = 15, l=-10)),
#     #legend.title = element_text(size = 10, margin = margin(b = 5, l=15)),
#     legend.title.position = 'top',
#     legend.key.size = unit(1.1, "lines"),
#     )
# print(p5)
#ggsave(path = '../experiments/figures/', 'fig5_actual_x_predicted_protGym_v03.2.png', width = 5, height = 3.5, units = 'in', dpi = 600)
```





# Supplementary Figures

## Supplementary 1

## Supplementary 2

## Supplementary 3

## Supplementary 4

```{r}
set.seed(55)

# make axes with ellipses
aa <- c("A", "C", "...", "... ", "W", "Y")
pos_labels <- c("1", "2", "3", "...", " ", "  ", "... ", "L-2", "L-1", "L")

n_aa <- length(aa)
L <- length(pos_labels)

# grid
df <- expand.grid(
  pos = factor(pos_labels, levels = pos_labels),
  aa  = factor(aa, levels = aa)
)

# dataframe for pooled
test_frac <- 0.2
df_rand <- df %>%
  mutate(
    split = ifelse(runif(n()) < test_frac, "test data", "train data"),
    scheme = "Pooled"
  )

# dataframe for stratified by site
test_pos <- sample(pos_labels, size = 2)
df_pos <- df %>%
  mutate(
    split = ifelse(pos %in% test_pos, "test data", "train data"),
    scheme = "Stratified by site"
  )

df_all <- bind_rows(df_rand, df_pos)

# plot
s4 <- ggplot(df_all, aes(x = pos, y = aa, fill = split)) +
  geom_tile(color = "black", linewidth = 0.1, linetype = 'solid') +
  scale_fill_manual(values = c("train data" = "#e1f7d5", "test data" = "#f1cbff")) +
  facet_wrap(~ scheme, nrow = 1) +
  labs(
    x = "Sequence position index",
    y = "Amino acid substitution"
  ) +
  theme_bw() +
  theme(
    legend.position = 'top',
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_text(angle = 90, vjust = 0.5, hjust = 0.5),
    legend.title = element_blank()
  )

print(s4)

ggsave(path = '../experiments/figures/', 'supp1_pooled_site_schematic_v02.png', width = 8, height = 4, units = 'in', dpi = 600, bg='white')
```

## Supplementary 5

## Supplementary 6

## Supplementary 7

## did we stil want this as a supplementary figure? pooled v stratified by site performance (Spearman)

```{r}
p4supp <- ggplot(full_res_avg, aes(x = Model, y = rho_score_test, fill = Split)) +
  scale_fill_manual(values = c("Pooled Split" = "#20B15C", "Site Split" = "#E2B40E"),
                    labels = c("Pooled", "Stratified by Site")) +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3, width = 0.5, alpha = 0.5) +
  geom_vline(xintercept = 2.5, linetype = 'dashed', color = 'gray', linewidth = 0.5) +
  scale_x_discrete(labels = c("ESM-2 650M", "ESM C 600M", "ESM-2 650M CRVDB", "ESM-2 650M URVDB", "ESM-2 3B Sawhney")) +
  labs(x = 'Model', y = expression(rho)) +
  theme_bw() +
  facet_wrap(~Source, nrow = 2) +
  theme(panel.grid.major.x = element_blank(), legend.position = 'top', legend.title = element_blank())

print(p4supp)

#ggsave('supp4_site_x_pooled_spearman.png', width = 8, height = 5, units = 'in', dpi = 300, path = '../experiments')
```



















